---
title: "Example : emhawkes package "
author: "Kyungsub Lee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Examle : emhawkes package}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Basic Hawkes model

### Univariate Hawkes process

```{r}
library(emhawkes)
```

This subsection outlines the steps for constructing, running simulations, and estimating a univariate Hawkes model.
To begin, create an `hspec` object which defines the Hawkes model.
S4 class `hspec` contains slots for the model parameters: `mu`, `alpha`, `beta`, `dimens`, `rmark`, and `impact`.

In a univariate model, the basic parameters of the model, `mu`, `alpha`, `beta`, can be given as numeric.
If numeric values are given, they will be converted to matrices.
Below is an example of a univariate Hawkes model without a mark.

```{r}
set.seed(1107)
mu1 <- 0.3; alpha1 <- 1.2; beta1 <- 1.5
hspec1 <- new("hspec", mu = mu1, alpha = alpha1, beta = beta1)
show(hspec1)
```

The function `hsim` implements simulation where the input arguments are `hspec`, `size` and the initial values of intensity component process, `lambda_component0`, and the initial values of Hawkes processes, `N0`.
More precisely, the intensity process the basic univariate Hawkes model is represented by $$ \lambda(t) = \mu + \int_{-\infty}^t \alpha e^{-\beta (t-s)} d N(s) = \mu + \lambda_c(0) e^{-\beta t} + \int_0^t \alpha e^{-\beta (t-s)} d N(s) $$ where the `lambda_component0` denotes $$ \lambda_c(0) = \int_{-\infty}^0 \alpha e^{\beta s} d N(s).$$ If `lambda_component0` is not provided, the internally determined initial values for intensity process are used.
If `size` is sufficiently large, exact value of `lambda_component0` may not be important.
The default initial value of counting process, `N0`, is zero.

```{r, warning=FALSE}
res1 <- hsim(hspec1, size = 100)
summary(res1)
```

The results of `hsim` is an S3 class `hreal` which consists of `hspec`, `inter_arrival`, `arrival`, `type`, `mark`, `N`, `Nc`, `lambda`, `lambda_component`, `rambda`, `rambda_component`.

-   `hspec` is the model specification

-   `inter_arrival` is the inter-arrival time of every event

-   `arrival` is the cumulative sum of `inter_arrival`

-   `type` is the type of events, i.e., $i$ for $N_i$, and used for multivariate model

-   `mark` is a numeric vector which represents additional information for the event

-   `lambda` represents $\lambda$ which is the left continuous and right limit version

-   The right continuous version of intensity is `rambda`

-   `lambda_component` represents $\lambda_{ij}$ and `rambda_component` is the right continuous version.

`inter_arrival`, `type`, `mark`, `N`, and `Nc` start at zero.
Using `summary()` function, one can print the first 20 elements of `arrival`, `N` and `lambda`.
`print()` function also can be used.

By the definition, we have `lambda == mu + lambda_compoent`:

```{r}
# first and third columns are the same
cbind(res1$lambda[1:5], res1$lambda_component[1:5], mu1 + res1$lambda_component[1:5])
```

Except the first row, `rambda == lambda + alpha` .

```{r}
# second and third columns are the same
cbind(res1$lambda[1:5], res1$rambda[1:5], res1$lambda[1:5] + alpha1)
```

Also check the exponential decaying:

```{r}
# By definition, the following two are equal:
res1$lambda[2:6]
mu1 + (res1$rambda[1:5] - mu1) * exp(-beta1 * res1$inter_arrival[2:6])
```

The log-likelihood function is computed by `logLik` method.
In this case, the inter-arrival times and `hspec` are inputs of the function.

```{r, warning=FALSE}
logLik(hspec1, inter_arrival = res1$inter_arrival)
```

The likelihood estimation is performed using `mhfit` function.
The specification of the initial values of the parameters, `hspec0` is needed.
Note that only `inter_arrival` is needed in this univariate model.
(Indeed, for more precise simulation, `lambda0`, the initial value of lambda component, should be specified. If not, internally determined initial values are used.) By default, it uses the BFGS method for numerical optimization.

```{r, warning=FALSE}
# initial value for numerical optimization
mu0 <- 0.5; alpha0 <- 1.0; beta0 <- 1.8
hspec0 <- new("hspec", mu = mu0, alpha = alpha0, beta = beta0)
# the intial values are provided through hspec
mle <- hfit(hspec0, inter_arrival = res1$inter_arrival)
summary(mle)
```

### Bivariate Hawkes model

The intensity process of basic bivariate Hawkes model is defined by

$$
 \lambda_1(t) = \mu_1 + \int_{-\infty}^t \alpha_{11} e^{-\beta_{11}(t-s)} d N_1(s) + \int_{-\infty}^t \alpha_{12} e^{-\beta_{12}(t-s)} d N_2(s),
$$

$$
 \lambda_2(t) = \mu_2 + \int_{-\infty}^t \alpha_{21} e^{-\beta_{21}(t-s)} d N_1(s) + \int_{-\infty}^t \alpha_{22} e^{-\beta_{22}(t-s)} d N_2(s).
$$

In a bivariate model, the parameters, the slots of `hspec`, are matrices.
`mu` is 2-by-1, and `alpha` and `beta` are 2-by-2 matrices:

$$
\mu = \begin{bmatrix} \mu_1 \\ \mu_2 \end{bmatrix}, \quad
\alpha = \begin{bmatrix} \alpha_{11} & \alpha_{12} \\ \alpha_{21} & \alpha_{22} \end{bmatrix}, \quad
\beta = 
\begin{bmatrix} \beta_{11} & \beta_{12} \\ \beta_{21} & \beta_{22} \end{bmatrix}
$$

`rmark` is a random number generating function for mark and is not used for non-mark model.
`lambda_component0`, 2-by-2 matrix, represents the initial values of `lambda_component`, a set of `lambda11, lambda12, lambda21, lambda22`.
The intensity processes are represented by

$$ \lambda_1(t) = \mu_1 + \lambda_{11}(t) + \lambda_{12}(t), $$

$$ \lambda_2(t) = \mu_2 + \lambda_{21}(t) + \lambda_{22}(t). $$

$\lambda_{ij}$ called lambda components and `lambda0` is $\lambda_{ij}(0)$.

`lambda_component0` can be omitted and then internally determined initial values are used.

```{r}
mu2 <- matrix(c(0.2), nrow = 2)
alpha2 <- matrix(c(0.5, 0.9, 0.9, 0.5), nrow = 2, byrow = TRUE)
beta2 <- matrix(c(2.25, 2.25, 2.25, 2.25), nrow = 2, byrow = TRUE)
hspec2 <- new("hspec", mu=mu2, alpha=alpha2, beta=beta2)
print(hspec2)
```

To simulate, use function `hsim`.

```{r}
res2 <- hsim(hspec2,  size=1000)
summary(res2)
```

`type` is crucial in multi-variate models, which represents the type of event.

```{r}
# Under bi-variate model, there are two types, 1 or 2.
res2$type[1:10]
```

The column names of `N` are `N1`, `N2`, `N3` and so on, for multivariate models.

```{r}
res2$N[1:3, ]
```

Similarly, the column names of `lambda` are `lambda1`, `lambda2`, `lambda3` and so on.

```{r}
res2$lambda[1:3, ]
```

The column names of `lambda_component` are `lambda_component11`, `lambda_component12`, `lambda_component13` and so on.

```{r}
res2$lambda_component[1:3, ]
```

By definition, the following two are the same:

```{r}
mu2[1] + rowSums(res2$lambda_component[1:5, c("lambda11", "lambda12")])
res2$lambda[1:5, "lambda1"]
```

From the result, we get vectors of realized `inter_arrival` and `type`.
Bivariate model requires `inter_arrival` and `type` for estimation.

```{r}
inter_arrival2 <- res2$inter_arrival
type2 <- res2$type
```

Log-likelihood is computed by a function `logLik`.

```{r}
logLik(hspec2, inter_arrival = inter_arrival2, type = type2)
```

A maximum log-likelihood estimation is performed using `hfit`.
In the following, the values of parameter slots in `hspec0`, such as `mu, alpha, beta`, serve as starting points of the numerical optimization.
For the purpose of illustration, we use `hspec0 <- hspec2`.
Since the true parameter values are not known in practical applications, the initial guess is used.
The realized `inter_arrival` and `type` are used for estimation.

```{r, warning=FALSE}
hspec0 <- hspec2
mle <- hfit(hspec0, inter_arrival = inter_arrival2, type = type2)
summary(mle)
```

```{r}
coef(mle)
```

```{r}
miscTools::stdEr(mle)
```

### Parameter setting

This subsection covers about the relation between parameter setting and estimation procedure in multi-variate Hawkes model.
The number of parameters to be estimated in the model depends on how we set the parameter slots such as `mu`, `alpha` and `beta` in `hspec0`, the specification for initial values.
Since the parameter slot such as `alpha` is a matrix, and the element in the matrix can be the same or different.
The number of parameters in the estimation varies depending on whether or not some of the elements in the initial setting are the same or different.

For example, if `alpha[1,1]` and `alpha[1,2]` in `hspec0` are different in initial starting, the numerical procedure tries to estimate both parameters of `alpha[1,1]` and `alpha[1,2]` differently.
If `alpha[1,1]` and `alpha[1,2]` are the same in the initial setting, then the estimation procedure considered two parameters are identical in the model and hence only one value is estimated.

Recall that the example in the previous section is of a symmetric Hawkes model where the matrix `alpha` is symmetric.
In addition, the elements of `beta` are all the same.

```{r}
print(hspec2)
```

```{r, warning=FALSE}
res2 <- hsim(hspec2, size = 1000)
```

In the first example of estimation, the initial value of `alpha0` is a matrix where the all elements have the same value of 0.75.
In this configuration, `hfit` assumes that `alpha11 == alpha12 == alpha21 == alpha22` in the model (even if the actual parameters have different values).
Similarly, the other parameter matrices `mu0` and `beta0` are also treated in the same way.

```{r, warning=FALSE}
mu0 <- matrix(c(0.15, 0.15), nrow = 2)
alpha0 <- matrix(c(0.75, 0.75, 0.75, 0.75), nrow = 2, byrow=TRUE)
beta0 <- matrix(c(2.6, 2.6, 2.6, 2.6), nrow = 2, byrow=TRUE)

hspec0 <- new("hspec", mu=mu0, alpha=alpha0, beta=beta0)
summary(hfit(hspec0, inter_arrival = res2$inter_arrival, type = res2$type))
```

Note that in the above result, `alpha1.1` is somewhere between original `alpha1.1 = 0.5` and `alpha1.2 = 0.9`.

In the following second example, `alpha0`'s elements are not same, but symmetric as in the original values in the simulation.
We have `alpha11 == alpha22` and `alpha11 == alpha22` in `alpha0` and hence `alpha11` and `alpha12` will be estimated differently.

```{r, warning=FALSE}
mu0 <- matrix(c(0.15, 0.15), nrow = 2)
alpha0 <- matrix(c(0.75, 0.751, 0.751, 0.75), nrow = 2, byrow=TRUE)
beta0 <- matrix(c(2.6, 2.6, 2.6, 2.6), nrow = 2, byrow=TRUE)

hspec0 <- new("hspec", mu=mu0, alpha=alpha0, beta=beta0)
summary(hfit(hspec0, inter_arrival = res2$inter_arrival, type = res2$type))
```

In the third example, the estimation is performed under the assumption that `mu1` and `mu2` may also be different (even though they are the same in the original model).

```{r, warning=FALSE}
mu0 <- matrix(c(0.15, 0.14), nrow = 2)
alpha0 <- matrix(c(0.75, 0.751, 0.751, 0.75), nrow = 2, byrow=TRUE)
beta0 <- matrix(c(2.6, 2.6, 2.6, 2.6), nrow = 2, byrow=TRUE)

hspec0 <- new("hspec", mu=mu0, alpha=alpha0, beta=beta0)
summary(hfit(hspec0, inter_arrival = res2$inter_arrival, type = res2$type))
```

By simply setting `reduced = FALSE`, all parameters are estimated (not recommended).

```{r, warning=FALSE}
summary(hfit(hspec2, inter_arrival = res2$inter_arrival, type = res2$type, reduced=FALSE))
```

The same logic is applied to all the higher dimensional model.

### Residual process

Residual process can be extracted by `residual_process()`.
`inter_arrival`, `type`, `rambda_component`, `mu`,`beta` should be provided.
The `component` denotes the type of the process to be extracted for multivariate model.
For example, for a bi-variate model, we have $N_1$ and \$N_2\$.
`component=1` is for the residual of $N_1$ and `component=2` is for the residual of $N_2$.

```{r, warning=FALSE, message = FALSE, fig.height=5, fig.width=5}
hrp <- new("hspec", mu = 0.3, alpha = 1.2, beta = 1.5)
res_rp <- hsim(hrp, size = 1000)

rp <- residual_process(component = 1,
                       inter_arrival = res_rp$inter_arrival, type = res_rp$type, 
                       rambda_component = res_rp$rambda_component, 
                       mu = 0.3, beta = 1.5)
p <- ppoints(100)
q <- quantile(rp,p=p)
plot(qexp(p), q, xlab="Theoretical Quantiles",ylab="Sample Quantiles")
qqline(q, distribution=qexp,col="blue", lty=2)
```

In case that `rambda_component` is unknown, it can be inferred by `infer_lambda()`.
For `infer_lambda()`, the model `hspec`, `inter_arrival` and `type` are required.
The above example is then:

```{r, message=FALSE, fig.height=5, fig.width=5}

# estimation
mle_rp <- hfit(new("hspec", mu = 0.2, alpha = 1, beta = 2),
               res_rp$inter_arrival)

# construct hspec from estimation result
he <- new("hspec", mu = coef(mle_rp)["mu1"], 
          alpha = coef(mle_rp)["alpha1"], beta = coef(mle_rp)["beta1"])

# infer intensity
infered_res <- infer_lambda(he, res_rp$inter_arrival, res_rp$type)

# compute residuals where we use 
rp2 <- residual_process(component = 1,
                       inter_arrival = res_rp$inter_arrival, type = res_rp$type, 
                       rambda_component = infered_res$rambda_component, 
                       mu = coef(mle_rp)["mu1"], beta = coef(mle_rp)["beta1"])
p <- ppoints(100)
q <- quantile(rp2, p=p)
plot(qexp(p), q, xlab="Theoretical Quantiles",ylab="Sample Quantiles")
qqline(q, distribution=qexp,col="blue", lty=2)
```

## More complicated model

### Multi-kernel model

In a multi-kernel Hawkes model, `type_col_map` is required for `hspec`.
`type_col_map` is a list that represents the mapping between type and column number.
For example, consider a bi-variate multi-kernel model: $$
\lambda_t = \mu + \int_{-\infty}^{t} h(t-u) d N(u)
$$ where $$ h = \sum_{k=1}^{K} h_k, \quad 
h_k (t) = \alpha_k \circ \begin{bmatrix}
e^{-\beta_{k11} t} & e^{-\beta_{k12} t} \\
e^{-\beta_{k21} t} & e^{-\beta_{k22} t}
\end{bmatrix}
 $$

with matrix $\alpha_k$ and $k$ denoting kernel number.

For example, in a bi-variate Hawkes model with two kernels, the intensity processes are

$$
 \begin{bmatrix} \lambda_1(t) \\ \lambda_2(t) \end{bmatrix} = 
\begin{bmatrix} \mu_1 \\ \mu_2 \end{bmatrix} + \int_{-\infty}^{t} \begin{bmatrix}
\alpha_{111} e^{-\beta_{111} t} & \alpha_{112} e^{-\beta_{112} t} \\
\alpha_{121}e^{-\beta_{121} t} & \alpha_{122}e^{-\beta_{122} t}
\end{bmatrix} \begin{bmatrix} d N_1(s) \\ dN_2(s) \end{bmatrix} 
+ \int_{-\infty}^{t} \begin{bmatrix}
\alpha_{211} e^{-\beta_{211} t} & \alpha_{212} e^{-\beta_{212} t} \\
\alpha_{221}e^{-\beta_{221} t} & \alpha_{222}e^{-\beta_{222} t}
\end{bmatrix} \begin{bmatrix} d N_1(s) \\ dN_2(s) \end{bmatrix}.
$$

The parameter matrix is defined by

$$
 \alpha = \begin{bmatrix} \alpha_{111} & \alpha_{112} & \alpha_{211} & \alpha_{212} \\ \alpha_{121} & \alpha_{122} & \alpha_{221} & \alpha_{222}
\end{bmatrix}, \quad
 \beta = \begin{bmatrix} \beta_{111} & \beta_{112} & \beta_{211} & \beta_{212} \\ \beta_{121} & \beta_{122} & \beta_{221} & \beta_{222}
\end{bmatrix} \quad
$$

and we should specify which columns of matrix are associated with which $N_i$.

```{r}
mu <- matrix(c(0.02, 0.02), nrow=2)
      
beta_1 <- matrix(rep(10, 4), nrow=2) 
beta_2 <- matrix(rep(1, 4), nrow=2)
beta  <- cbind(beta_1, beta_2)
      
alpha_1 <- matrix(c(3, 2,
                    2, 3), nrow=2, byrow=TRUE)
alpha_2 <- matrix(c(0.3, 0.2,
                    0.2, 0.3), nrow=2, byrow=TRUE)
alpha <- cbind(alpha_1, alpha_2)

print(alpha)
```

Note that $d N_1(s)$ is multiplied by first and third columns of $\alpha$ and $dN_2(s)$ is multiplied by second and fourth columns of $\alpha$ and hence `type_col_map` is

```{r}
type_col_map <- list(c(1,3),  # columns for dN1
                     c(2,4))  # columns for dN2
type_col_map
```

where type `i` is associated with columns of `type_col_map[[i]]`.
Thus,

```{r}
cat("Part of alpha associated with N1: \n")
alpha[, type_col_map[[1]]]  # associated with N1
cat("Part of alpha associated with N2: \n")
alpha[, type_col_map[[2]]]  # associated with N2

cat("Part of beta associated with N1: \n")
beta[, type_col_map[[1]]]  # associated with N1
cat("Part of beta associated with N2: \n")
beta[, type_col_map[[2]]]  # associated with N2
```

```{r}
h <- new("hspec", mu = mu, alpha = alpha, beta=beta, type_col_map = type_col_map)
h
```

In addition, `lambda_component0` should be provided for simulation and estimation.

```{r}

res_mk <- hsim(h, size = 2000, 
               # for an illustration purpose
               lambda_component0 = matrix(seq(1, 1.7, 0.1), nrow = 2)) 
res_mk
```

```{r, warning=FALSE}
summary(hfit(h, res_mk$inter_arrival, res_mk$type,
             lambda_component0 = matrix(seq(1, 1.7, 0.1), nrow = 2)))
```

### Linear impact model

Marked Hawkes models are represented by

$$
 \lambda(t) = \mu + \int_{(-\infty, t) \times E} (\alpha + \Psi(z))e^{-\beta(t-u)} M(du \times dz).
$$

In the following, `impact` represents $\Psi(z)$, the impact of mark on future intensity.
It is a function, and the first argument is `param` represents the parameter of the model.
`impact` function can have additional arguments such as `alpha`, `n`, `mark`, etc., which represents the dependence.
Do not miss `...` as the ellipsis is omitted, an error occurs.
`rmark()` is a function that generate marks for simulation.

```{r, warning=FALSE}
mu <- matrix(c(0.15, 0.15), nrow=2)
alpha <- matrix(c(0.75, 0.6, 0.6, 0.75), nrow=2, byrow=T)
beta <- matrix(c(2.6, 2.6, 2.6, 2.6), nrow=2)
rmark <- function(param = c(p=0.65), ...){
  rgeom(1, p=param[1]) + 1
}

impact <- function(param = c(eta1=0.2), alpha, n, mark, ...){
  ma <- matrix(rep(mark[n]-1, 4), nrow = 2)
  alpha * ma * matrix( rep(param["eta1"], 4), nrow=2)
}

hi <- new("hspec", mu=mu, alpha=alpha, beta=beta,
          rmark = rmark,
          impact=impact)
hi
```

```{r}

res <- hsim(hi, size=1000, lambda_component0 = matrix(rep(0.1,4), nrow=2))

fit <- hfit(hi, 
            inter_arrival = res$inter_arrival,
            type = res$type,
            mark = res$mark,
            lambda_component0 = matrix(rep(0.1,4), nrow=2))

summary(fit)
```

For a special case of linear impact function, the following implementation is recommended.
In a marked Hawkes model, the additional linear impact can be represented by slot `eta`.
In this model, the intensity process is

$$
 \lambda(t) = \mu + \int_{(-\infty, t)\times E} (\alpha + \eta (z-1)) e^{-\beta(t-u)}  M(dt \times dz).
$$

where $E$ is a mark space.

```{r}

rmark <- function(param = c(p=0.65), ...){
  rgeom(1, p=param[1]) + 1
}

h <-  new("hspec", mu=0.15, alpha=0.7, beta=1.6, eta=0.3,
          rmark = rmark)
h
```

```{r}
res <- hsim(h, size = 1000)
summary(res)
```

```{r}
fit <- hfit(h, 
            inter_arrival = res$inter_arrival,
            type = res$type,
            mark = res$mark)
summary(fit)
```

### Hawkes flocking model

In the basic model, `alpha` is a matrix, but it can be a function as in the following code.
The function `alpha` simply return a $4\times4$ matrix but by doing so, we can fix some of elements as specific vales when estimating.
When estimating, the optimization only works for the specified parameters in `param`.
In the case of simulation, there is no difference whether the parameter set is represented by a matrix or a function.

```{r, warning=FALSE}
mu <- matrix(c(0.02, 0.02, 0.04, 0.04), nrow = 4)


alpha <- function(param = c(alpha11 = 0, alpha12 = 0.3, alpha33 = 0.3, alpha34 = 0.4)){
  matrix(c(param["alpha11"], param["alpha12"], 0, 0,
           param["alpha12"], param["alpha11"], 0, 0,
           0, 0, param["alpha33"], param["alpha34"],
           0, 0, param["alpha34"], param["alpha33"]), nrow = 4, byrow = T)
}


beta <- matrix(c(rep(0.7, 8), rep(1.1, 8)), nrow = 4, byrow = T)

impact <- function(param = c(alpha1n=0, alpha1w=0.1, alpha2n=0.1, alpha2w=0.2),
                   n=n, N=N, ...){
  
  Psi <- matrix(c(0, 0, param['alpha1w'], param['alpha1n'],
                  0, 0, param['alpha1n'], param['alpha1w'],
                  param['alpha2w'], param['alpha2n'], 0, 0,
                  param['alpha2n'], param['alpha2w'], 0, 0), nrow=4, byrow=T)
  
  ind <- N[,"N1"][n] - N[,"N2"][n] > N[,"N3"][n] - N[,"N4"][n] + 0.5
  
  km <- matrix(c(!ind, !ind, !ind, !ind,
                 ind, ind, ind, ind,
                 ind, ind, ind, ind,
                 !ind, !ind, !ind, !ind), nrow = 4, byrow = T)
  
  km * Psi
}

h <- new("hspec",
         mu = mu, alpha = alpha, beta = beta, impact = impact)

hr <- hsim(h, size=1000)


fit <- hfit(h, hr$inter_arrival, hr$type)
summary(fit)
```

### Flocking in intensity model

```{r}
mu <- function(param = c(theta_p = 0.2, theta_n = 0.2, kappa = 0.1)){
  theta    <- (param["theta_p"] + param["theta_n"])/2
  theta_tl <- (param["theta_n"] - param["theta_n"])/2
  matrix(c(theta/2/(1 - param["kappa"]) + theta_tl/2/(1 + param["kappa"]),
           theta/2/(1 - param["kappa"]) - theta_tl/2/(1 + param["kappa"])), nrow=2)
}

alpha <- function(param = c(eta = 10, nu = 9)){
  zeta    <- (param["eta"] + param["nu"])/2
  zeta_tl <- (param["eta"] - param["nu"])/2
  matrix(c(zeta, zeta_tl, zeta, -zeta_tl,
           zeta, -zeta_tl, zeta, zeta_tl), nrow=2, byrow=TRUE)
}

beta <- function(param = c(kappa = 0.1, beta = 30)){
  beta1 <- param["beta"] * (1 - param["kappa"])
  beta2 <- param["beta"] * (1 + param["kappa"])
  matrix(c(beta1, beta2, beta1, beta2,
           beta1, beta2, beta1, beta2), nrow = 2, byrow = TRUE)
}

type_col_map <- list(c(1,2), c(3,4))

h <- new("hspec", mu = mu, alpha = alpha, beta = beta, type_col_map = type_col_map)

# run simulation
res <- hsim(h, size = 1000, lambda_component0 = matrix(rep(1, 2 * 4), nrow=2))
summary(res)


```

### Bid-ask price model

```{r, warning=FALSE}

initial_ask_price <- 1200.5
initial_bid_price <- 1199.5
scaling <- 100
initial_level <- round((initial_ask_price - initial_bid_price)*scaling - 1)
initial_mid_price <- (initial_bid_price + initial_ask_price)/2


mu <- function(param = c(mu1 = 0.08, zeta1 = 0.10), n=n, Nc=Nc, ...){

  if(n == 1){
    level <- initial_level
    mid <- initial_mid_price
  } else {
    level <- Nc[n-1,1] - Nc[n-1,2] - (Nc[n-1,3] - Nc[n-1,4]) + initial_level
    ask <- initial_ask_price + (Nc[n-1,1] - Nc[n-1,2])/100
    bid <- initial_bid_price + (Nc[n-1,3] - Nc[n-1,4])/100
    mid <- (ask + bid)/2
  }
  if(level <= 0){
    matrix(c(param["mu1"], 0,
             0, param["mu1"]), nrow = 4)
  } else {
    matrix(c(param["mu1"], param["zeta1"][[1]]*level/mid*scaling,
             param["zeta1"][[1]]*level/mid*scaling, param["mu1"]), nrow = 4)

  }

}

alpha <- function(param = c(alpha_s1=4, alpha_m=26, alpha_s2=5,
                            alpha_w1=11, alpha_w2=7)){
  matrix(c(param["alpha_s1"], param["alpha_m"], param["alpha_s2"], 0,
           param["alpha_w1"], 0, 0, param["alpha_w2"],
           param["alpha_w2"], 0, 0, param["alpha_w1"],
           0, param["alpha_s2"], param["alpha_m"], param["alpha_s1"]), nrow = 4, byrow = TRUE)
}

impact <- function(param = c(xi = 2.7), n=n, Nc=Nc, lambda_component_n = lambda_component_n, ... ){
  if(n == 1){
    level <-  initial_level
    mid <- initial_mid_price
  } else {
    level <- Nc[n,1] - Nc[n,2] - (Nc[n,3] - Nc[n,4]) + initial_level
    ask <- initial_ask_price + (Nc[n,1] - Nc[n,2])/100
    bid <- initial_bid_price + (Nc[n,3] - Nc[n,4])/100
    mid <- (ask + bid)/2
  }

  l2 <- sum(lambda_component_n[2,])
  l3 <- sum(lambda_component_n[3,])

  im <- matrix(c(0, 0, 0, 0,
                 0, -l2 + param["xi"]*level/mid*scaling, -l2 + param["xi"]*level/mid*scaling, 0,
                 0, -l3 + param["xi"]*level/mid*scaling, -l3 + param["xi"]*level/mid*scaling, 0,
                 0, 0, 0, 0), nrow = 4, byrow = TRUE)

}
beta <- matrix(rep(50, 16), nrow = 4, byrow=TRUE)

rmark <- function(param = c(p=0.3), n=n, Nc=Nc, type, ...){
  if(n == 1){
    level <-  initial_level
  } else {
    level <- Nc[n-1,1] - Nc[n-1,2] - (Nc[n-1,3] - Nc[n-1,4]) + initial_level
  }
  if (type[n] == 2 | type[n] == 3){
    min(level,  rgeom(1, p=0.65) + 1)
  } else {
    rgeom(1, p=0.65) + 1
  }
}



h <- new("hspec", mu = mu, alpha = alpha, beta = beta, impact=impact, rmark = rmark)

hr <- hsim(h, size=1001, lambda_component0 = matrix(rep(1, 16), 4))
summary(hr)
```
